---
title: "Degrees of Freedom Adjustment in hetid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Degrees of Freedom Adjustment in hetid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hetid)
```

## Introduction

The `hetid` package now supports both asymptotic and finite sample standard errors through the `df_adjust` parameter. This vignette explains the difference between these approaches and how to use them.

## Background

When computing standard errors in regression models, there are two main approaches:

1. **Asymptotic standard errors**: Use the formula $SE = \sqrt{\hat{\sigma}^2 (X'X)^{-1}}$ where $\hat{\sigma}^2 = RSS/n$
2. **Finite sample standard errors**: Use $\hat{\sigma}^2 = RSS/(n-k)$ where $k$ is the number of parameters

The difference is in the degrees of freedom adjustment. Asymptotic theory assumes $n \to \infty$, while finite sample corrections account for the loss of degrees of freedom from estimating parameters.

## Default Behavior

Following Stata's convention, **asymptotic standard errors are now the default** in hetid:

```{r default-example}
# Generate example data
params <- list(
  beta1_0 = 0.5, beta1_1 = 1.5, gamma1 = -0.8,
  beta2_0 = 1.0, beta2_1 = -1.0,
  alpha1 = -0.5, alpha2 = 1.0, delta_het = 1.2,
  sample_size = 200,
  tau_set_id = 0,
  bootstrap_reps = 50
)

# Run with default (asymptotic) standard errors
result_asymp <- run_single_lewbel_simulation(
  sim_id = 1,
  params = params,
  df_adjust = "asymptotic" # This is the default
)

# Display results
cat("Asymptotic SEs:\n")
cat("OLS SE:", result_asymp$ols_se, "\n")
cat("TSLS SE:", result_asymp$tsls_se, "\n")
```

## Using Finite Sample Standard Errors

For smaller samples, you may want to use finite sample corrections:

```{r finite-example}
# Run with finite sample standard errors
result_finite <- run_single_lewbel_simulation(
  sim_id = 1,
  params = params,
  df_adjust = "finite"
)

# Display results
cat("\nFinite Sample SEs:\n")
cat("OLS SE:", result_finite$ols_se, "\n")
cat("TSLS SE:", result_finite$tsls_se, "\n")

# Calculate the adjustment factor
n <- params$sample_size
k <- 3 # intercept + 2 regressors
adjustment_factor <- sqrt(n / (n - k))
cat("\nAdjustment factor:", adjustment_factor, "\n")
cat(
  "Ratio finite/asymptotic:",
  result_finite$tsls_se / result_asymp$tsls_se, "\n"
)
```

## Impact on Confidence Intervals

The `df_adjust` parameter affects both standard errors and critical values:

```{r confidence-intervals}
# Critical values
n <- 200
k <- 3
crit_asymp <- get_critical_value(n, k, alpha = 0.05, df_adjust = "asymptotic")
crit_finite <- get_critical_value(n, k, alpha = 0.05, df_adjust = "finite")

cat("Critical values for 95% CI:\n")
cat("Asymptotic (normal):", crit_asymp, "\n")
cat("Finite (t-distribution):", crit_finite, "\n")
cat("Difference:", (crit_finite - crit_asymp) / crit_asymp * 100, "%\n")
```

## Monte Carlo Simulations

When running Monte Carlo simulations, you can specify the df_adjust in the configuration:

```{r monte-carlo, eval=FALSE}
# Create configuration with finite sample adjustment
config <- create_default_config()
config$df_adjust <- "finite"

# Run simulation
results <- run_lewbel_monte_carlo(config)

# Compare with asymptotic
config_asymp <- create_default_config()
config_asymp$df_adjust <- "asymptotic"
results_asymp <- run_lewbel_monte_carlo(config_asymp)
```

## Recommendations

1. **For large samples (n > 200)**: The difference between asymptotic and finite sample SEs is negligible
2. **For small samples (n < 100)**: Consider using finite sample corrections
3. **For comparison with Stata**: Use asymptotic (the default) as Stata's ivreg2h reports asymptotic SEs
4. **For comparison with R's lm()**: Use finite sample, as lm() uses n-k by default

## Technical Details

The adjustment is implemented through several helper functions:

- `adjust_se_for_df()`: Applies the degrees of freedom correction
- `get_critical_value()`: Returns appropriate critical values (normal vs t-distribution)
- `extract_se_lm()`: Extracts SEs from lm models with adjustment
- `extract_se_ivreg()`: Extracts SEs from ivreg models with adjustment

These functions ensure consistent behavior across all estimation methods in the package.
