---
title: "Semiparametric Control Function Approach: The Klein & Vella (2010) Method"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  pdf_document:
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Semiparametric Control Function Approach: The Klein & Vella (2010) Method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: https://www.zotero.org/styles/journal-of-econometrics
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(hetid)
library(ggplot2)
library(dplyr)
```

## Introduction

This vignette demonstrates the Klein & Vella (2010) method for identification through heteroskedasticity using a control function approach. Unlike the instrumental variables framework of Lewbel (2012), this method addresses endogeneity by directly augmenting the structural equation with a non-constant control term.

> **Related vignettes**: For theoretical background, see [Theory and Methods](heteroskedasticity-identification-theory.html). For IV-based approaches, see [Getting Started](getting-started.html) (Lewbel), [Rigobon Method](rigobon-method.html) (regime-based), or [Prono Method](prono-method.html) (GARCH-based).

### The Control Function Approach

The key insight of @klein2010estimating is that under certain conditions about the error structure, endogeneity can be controlled for by including a specific function of the first-stage residuals in the main equation. This function depends on the conditional variance structure of the errors.

### Klein & Vella vs Other Methods

Both Klein & Vella and Lewbel exploit heteroskedasticity for identification, but they differ fundamentally:

- **Lewbel (2012)**: Uses heteroskedasticity to construct instrumental variables
- **Klein & Vella (2010)**: Uses heteroskedasticity to construct a control function
- **Rigobon (2003)**: Uses discrete regime changes (can be adapted to either IV or control function)
- **Prono (2014)**: Uses time-varying conditional heteroskedasticity (GARCH)

## Theoretical Background

### The Model

Consider the triangular system:

$$\begin{aligned}
Y_1 &= X^\top\beta_1 + \gamma_1 Y_2 + \varepsilon_1 \\
Y_2 &= X^\top\beta_2 + \varepsilon_2
\end{aligned}$$

The general control function decomposition is:
$$\varepsilon_1 = A(X)\varepsilon_2 + \eta_1$$

where $A(X) = \frac{\text{Cov}(\varepsilon_1, \varepsilon_2 \mid X)}{\text{Var}(\varepsilon_2 \mid X)}$ and $E[\eta_1 | X, \varepsilon_2] = 0$.

### Key Assumptions

1. **Exogeneity**: $E[\varepsilon_j | X] = 0$ for $j=1,2$
2. **Constant Conditional Correlation**: $\text{Corr}(\varepsilon_1, \varepsilon_2 | X) = \rho_0$
3. **Heteroskedasticity and Variation**: The ratio $S_1(X)/S_2(X)$ is not constant

Under these assumptions, the control function simplifies to:
$$A(X) = \rho_0 \frac{S_1(X)}{S_2(X)}$$

where $S_j(X) = \sqrt{\text{Var}(\varepsilon_j|X)}$ are the conditional standard deviations.

## Basic Usage

### Running a Simple Demonstration

```{r klein-vella-demo}
# Run the Klein & Vella demonstration
run_klein_vella_demo(n = 500)
```

### Understanding the Output

The demo shows:
- **True parameters**: The actual values used to generate the data
- **Parametric estimates**: Results using exponential variance specification
- **Semiparametric estimates**: Results using nonparametric variance estimation (if `np` package available)
- **Comparison with OLS**: Shows the bias correction achieved

## Parametric Implementation

The parametric version assumes specific functional forms for the conditional variances.

### Exponential Variance Model

The most common specification uses exponential functions to ensure positivity:
$$S_j^2(X) = \exp(X^\top\delta_j)$$

```{r parametric-example}
# Generate data with heteroskedasticity suitable for Klein & Vella
set.seed(123)
n <- 1000

# Create configuration
config <- create_klein_vella_config(
  n = n,
  beta1 = c(0.5, 1.5), # Coefficients in first equation
  beta2 = c(1.0, -1.0), # Coefficients in second equation
  gamma1 = -0.8, # Endogenous parameter
  rho = 0.6, # Correlation between errors
  delta1 = c(0.1, 0.3), # Variance parameters for epsilon1
  delta2 = c(0.2, -0.2), # Variance parameters for epsilon2
  seed = 123
)

# Generate data
data <- generate_klein_vella_data(config)

# Examine the generated data
head(data)

# Check heteroskedasticity patterns
summary(data)
```

### Estimation

```{r parametric-estimation}
# Estimate using parametric Klein & Vella
kv_results <- klein_vella_parametric(
  data = data,
  y1_var = "Y1",
  y2_var = "Y2",
  x_vars = "X",
  variance_type = "exponential",
  verbose = TRUE
)

# Display results
print(kv_results)

# Compare with OLS (biased)
ols_model <- lm(Y1 ~ X + Y2, data = data)
cat("\nOLS estimate of gamma1:", coef(ols_model)["Y2"], "\n")
cat("Klein-Vella estimate:", kv_results$estimates["gamma1"], "\n")
cat("True value:", config$gamma1, "\n")
```

### Visualizing the Control Function

```{r viz-control-function, fig.alt="Scatter plot showing the relationship between X and the estimated control function, demonstrating how the control term varies with X due to heteroskedasticity"}
# Extract control function values
control_values <- kv_results$control_function

# Create visualization
plot_df <- data.frame(
  X = data$X,
  Control = control_values
)

ggplot(plot_df, aes(x = X, y = Control)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(
    title = "Klein & Vella Control Function",
    subtitle = expression(paste("Control term: ", rho[0], " × ", frac(S[1](X), S[2](X)), " × ", hat(epsilon)[2])),
    x = "Exogenous Variable (X)",
    y = "Control Function Value"
  ) +
  theme_minimal()
```

## Semiparametric Implementation

The semiparametric version estimates the variance functions nonparametrically.

```{r semiparametric-check}
# Check if np package is available
np_available <- requireNamespace("np", quietly = TRUE)

if (!np_available) {
  cat("Note: The 'np' package is not installed.\n")
  cat("For semiparametric estimation, install it with: install.packages('np')\n")
  cat("Proceeding with parametric estimation only.\n\n")
}
```

```{r semiparametric-estimation, eval=FALSE}
# This code requires the np package
if (np_available) {
  # Estimate using semiparametric Klein & Vella
  kv_semi_results <- klein_vella_semiparametric(
    data = data,
    y1_var = "Y1",
    y2_var = "Y2",
    x_vars = "X",
    bandwidth_method = "cv.aic", # Cross-validation with AIC
    verbose = TRUE
  )

  # Compare parametric vs semiparametric
  comparison <- data.frame(
    Method = c("OLS", "Parametric K&V", "Semiparametric K&V", "True Value"),
    gamma1 = c(
      coef(ols_model)["Y2"],
      kv_results$estimates["gamma1"],
      kv_semi_results$estimates["gamma1"],
      config$gamma1
    )
  )

  print(comparison)
}
```

## Comparison with Lewbel Method

Let's compare Klein & Vella with Lewbel on the same dataset.

```{r comparison-with-lewbel}
# Generate data that satisfies both Klein-Vella and Lewbel assumptions
comparison_data <- generate_klein_vella_data(config)

# Add Z variable for Lewbel
comparison_data$Z <- comparison_data$X^2 - mean(comparison_data$X^2)

# Klein & Vella estimate (from above)
kv_gamma1 <- kv_results$estimates["gamma1"]

# Lewbel 2SLS estimate
e2_lewbel <- residuals(lm(Y2 ~ X, data = comparison_data))
iv_lewbel <- comparison_data$Z * e2_lewbel
lewbel_model <- AER::ivreg(Y1 ~ X + Y2 | X + iv_lewbel, data = comparison_data)
lewbel_gamma1 <- coef(lewbel_model)["Y2"]

# Create comparison
method_comparison <- data.frame(
  Method = c("OLS (biased)", "Klein & Vella", "Lewbel 2SLS", "True Value"),
  Estimate = c(
    coef(ols_model)["Y2"],
    kv_gamma1,
    lewbel_gamma1,
    config$gamma1
  ),
  Bias = c(
    coef(ols_model)["Y2"] - config$gamma1,
    kv_gamma1 - config$gamma1,
    lewbel_gamma1 - config$gamma1,
    0
  )
)

print(method_comparison, digits = 4)

# Visualize comparison
ggplot(method_comparison, aes(x = Method, y = Estimate)) +
  geom_point(size = 4, color = "darkblue") +
  geom_hline(yintercept = config$gamma1, linetype = "dashed", color = "red") +
  labs(
    title = "Comparison of Identification Methods",
    subtitle = paste("True value =", config$gamma1, "(red line)"),
    y = "Estimate of γ₁"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Monte Carlo Simulation

To assess the finite sample properties of the Klein & Vella estimator:

```{r monte-carlo, eval=FALSE}
# Run Monte Carlo simulation
mc_results <- run_klein_vella_monte_carlo(
  config = config,
  n_sims = 500,
  methods = c("ols", "klein_vella_param", "lewbel"),
  parallel = FALSE,
  progress = TRUE
)

# Summary statistics
summary_stats <- mc_results %>%
  group_by(method) %>%
  summarise(
    mean_estimate = mean(gamma1_est),
    bias = mean(gamma1_est - config$gamma1),
    std_error = sd(gamma1_est),
    rmse = sqrt(mean((gamma1_est - config$gamma1)^2)),
    coverage_95 = mean(gamma1_est - 1.96 * gamma1_se <= config$gamma1 &
      gamma1_est + 1.96 * gamma1_se >= config$gamma1)
  )

print(summary_stats)

# Distribution plot
ggplot(mc_results, aes(x = gamma1_est, fill = method)) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = config$gamma1, linetype = "dashed") +
  facet_wrap(~method, scales = "free_y") +
  labs(
    title = "Monte Carlo Distribution of Estimates",
    subtitle = paste("True value =", config$gamma1, "(dashed line)"),
    x = "Estimate",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Advanced Features

### Multiple X Variables

The method extends naturally to multiple exogenous variables:

```{r multiple-x}
# Configuration with multiple X variables
config_multi <- create_klein_vella_config(
  n = 1000,
  k = 3, # Number of X variables
  beta1 = c(0.5, 1.0, -0.5, 0.8), # Intercept + 3 X coefficients
  beta2 = c(1.0, -0.5, 0.7, -0.3), # Intercept + 3 X coefficients
  gamma1 = -0.8,
  rho = 0.6,
  delta1 = c(0.1, 0.2, -0.1, 0.15), # Variance function parameters
  delta2 = c(0.2, -0.2, 0.1, -0.1)
)

# Generate and estimate
data_multi <- generate_klein_vella_data(config_multi)
results_multi <- klein_vella_parametric(
  data = data_multi,
  y1_var = "Y1",
  y2_var = "Y2",
  x_vars = c("X1", "X2", "X3")
)

print(results_multi)
```

### Testing Key Assumptions

```{r test-assumptions}
# Test constant correlation assumption
test_constant_correlation <- function(data, n_groups = 5) {
  # Divide data into groups based on X
  if ("X" %in% names(data)) {
    x_var <- data$X
  } else {
    x_var <- data$X1 # Use first X if multiple
  }

  data$group <- cut(x_var, breaks = n_groups)

  # Estimate correlations by group
  correlations <- data %>%
    group_by(group) %>%
    summarise(
      n = n(),
      correlation = cor(Y1 - mean(Y1), Y2 - mean(Y2))
    )

  # Test for equality
  cat("Correlations by X group:\n")
  print(correlations)

  # Simple F-test for constant correlation
  # In practice, use more sophisticated tests
  return(correlations)
}

# Apply test
test_constant_correlation(data)

# Test heteroskedasticity (should be present)
library(lmtest)
bptest(lm(Y2 ~ X, data = data))
```

## Practical Recommendations

### When to Use Klein & Vella

The Klein & Vella method is particularly suitable when:

1. **No valid instruments available**: Unlike Lewbel, doesn't require constructing IVs
2. **Constant correlation plausible**: The error correlation doesn't vary with X
3. **Clear heteroskedasticity**: The variance ratio $S_1(X)/S_2(X)$ varies substantially
4. **Triangular system**: Currently limited to recursive models

### Implementation Choices

1. **Parametric vs Semiparametric**:
   - Use parametric for speed and when variance structure is well-understood
   - Use semiparametric for robustness when unsure about functional forms

2. **Variance Specification** (parametric):
   - Exponential: Most common, ensures positivity
   - Power: $S_j^2(X) = (X^\top\delta_j)^2$ for interpretability
   - Custom: Any positive function can be used

3. **Bandwidth Selection** (semiparametric):
   - Cross-validation methods are most reliable
   - Rule-of-thumb can be faster for initial exploration

### Comparison with Other Methods

| Method | Approach | Key Assumption | Advantages | Limitations |
|--------|----------|----------------|------------|-------------|
| Klein & Vella | Control Function | Constant correlation | No instruments needed | Triangular only |
| Lewbel | IV/2SLS | Covariance restriction | Works for simultaneous | Needs instrument relevance |
| Rigobon | Regime-based | Discrete variance changes | Clear interpretation | Requires regime indicator |
| Prono | GARCH-based | Time-varying variance | Natural for time series | Requires long series |

## Diagnostics and Validation

### Checking the Variance Ratio

```{r variance-ratio-diagnostic, fig.alt="Scatter plot showing how the variance ratio S1/S2 changes with X, which is crucial for Klein-Vella identification"}
# Extract estimated variance functions
if (exists("kv_results") && !is.null(kv_results$variance_functions)) {
  s1_values <- sqrt(kv_results$variance_functions$S1_squared)
  s2_values <- sqrt(kv_results$variance_functions$S2_squared)

  ratio_df <- data.frame(
    X = data$X,
    Ratio = s1_values / s2_values
  )

  # Check variation in ratio
  cat("Variance ratio statistics:\n")
  cat("Mean:", mean(ratio_df$Ratio), "\n")
  cat("SD:", sd(ratio_df$Ratio), "\n")
  cat("CV:", sd(ratio_df$Ratio) / mean(ratio_df$Ratio), "\n")

  # Plot
  ggplot(ratio_df, aes(x = X, y = Ratio)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "loess", color = "red") +
    geom_hline(yintercept = mean(ratio_df$Ratio), linetype = "dashed") +
    labs(
      title = "Variance Ratio for Klein & Vella Identification",
      subtitle = "Variation in S₁(X)/S₂(X) is key for identification",
      x = "X",
      y = "S₁(X)/S₂(X)"
    ) +
    theme_minimal()
}
```

### Bootstrap Inference

```{r bootstrap, eval=FALSE}
# Bootstrap for more accurate inference
bootstrap_kv <- function(data, n_boot = 200) {
  boot_results <- replicate(n_boot, {
    # Resample data
    boot_indices <- sample(nrow(data), replace = TRUE)
    boot_data <- data[boot_indices, ]

    # Estimate
    boot_est <- klein_vella_parametric(
      data = boot_data,
      y1_var = "Y1",
      y2_var = "Y2",
      x_vars = "X",
      verbose = FALSE
    )

    boot_est$estimates["gamma1"]
  })

  # Confidence interval
  ci <- quantile(boot_results, c(0.025, 0.975))
  se <- sd(boot_results)

  list(
    mean = mean(boot_results),
    se = se,
    ci_lower = ci[1],
    ci_upper = ci[2]
  )
}

# Run bootstrap
boot_results <- bootstrap_kv(data)
cat("Bootstrap results for gamma1:\n")
cat("Estimate:", boot_results$mean, "\n")
cat("SE:", boot_results$se, "\n")
cat("95% CI:", boot_results$ci_lower, "to", boot_results$ci_upper, "\n")
```

## Extensions and Future Work

### Simultaneous Equations

While the current implementation focuses on triangular systems, the Klein & Vella approach could potentially be extended to simultaneous equations under additional assumptions.

### Time-Varying Parameters

For time series applications, the method could be combined with time-varying parameter models where the correlation structure evolves smoothly over time.

### Panel Data

Extension to panel data would require modeling both within-group and between-group heteroskedasticity patterns.

## Conclusion

The Klein & Vella (2010) method provides a powerful alternative to IV-based approaches for handling endogeneity through heteroskedasticity. Its control function framework is particularly valuable when:

- Traditional instruments are unavailable
- The constant correlation assumption is reasonable
- Heteroskedasticity patterns are strong

The `hetid` package provides both parametric and semiparametric implementations, allowing researchers to choose the approach that best fits their application.

## See Also

- [Theory and Methods](heteroskedasticity-identification-theory.html) - Mathematical foundations
- [Getting Started](getting-started.html) - Basic Lewbel (2012) implementation
- [Rigobon Method](rigobon-method.html) - Regime-based identification
- [Prono Method](prono-method.html) - GARCH-based identification
- [Package Comparison](package-comparison.html) - Software validation

## References
